---
title: "chi square for RNA types"
author: "Megan Ewing"
date: "2025-01-30"
output: html_document
---

```{r}


# only need to run install once

# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("cowplot")
# install.packages("ggpubr")
# install.packages("rstatix")
# install.packages("BiocManager")
# BiocManager::install("DESeq2")
# install.packages("patchwork")

# packages -- may need to load each time project is open 

  library(dplyr)
  library(tidyverse)
  library(tidyr)
  library(data.table)
  library(ggplot2)
  library(RColorBrewer)
  library(cowplot)
  library(ggpubr)
  library(rstatix)
  library(DESeq2)

```

```{r}
#read in the count table from STAR (or DESEq2)<-ask Giles
#note: in BBEdit I replaced the "Type" information: replaced "protein-coding" with "mito" for the first 13 genes 
#which are the mitochondrial genes

counts<- read.csv("../data/chisq_rnatypes/count_data_mitoedit.csv", row.names = 1)
head(counts)

```

### Transform Data to Proportions

doing this to account for different library lengths in our reads. counts will be replaced by proportion of reads. For example, M.C.216 "gene 1" will no longer be count #, but will be count \# divided by the total number of reads for that sample.

```{r}
# create new dataframe of proportions from counts
# can change this from counts to norm_counts to get proportion of normalized counts 
props <- norm_counts  

# Apply function to columns with count data (1-30) that will divide each cell by the total number of reads for that sample (ie. cell # / sum of col)
props[, 1:30] <- lapply(norm_counts[, 1:30], function(col) col / sum(col))  

head(props)

```

### Normalize Count Data Using DESeq2

preparation of count data

```{r}
# Read the FeatureCounts output file into R
featureCounts <- read.csv("../output/STAR_count_data.csv", header=TRUE, row.names=1)
head(featureCounts)

#subset from feature counts of just the count data
countData <- featureCounts[, -(31:33)] 

# adjusting for sample 193
#confirm location
colnames(countData[16])

#we're moving it to first postion
countData <- countData[, c(16, 1:15, 17:ncol(countData))]
head(countData)

# specifying metadata
metaData <- data.frame(
  sample = c("M.C.193", "M.C.216", "M.C.218", "M.C.226", "M.C.306", "M.C.329", "M.C.334", "M.C.337", "M.C.339", "M.C.358", "M.C.360", "M.C.363", "M.C.373", "M.C.482","M.C.488", "M.T.18",  "M.T.20", "M.T.22", "M.T.235", "M.T.245", "M.T.29", "M.T.31", "M.T.32", "M.T.399", "M.T.43", "M.T.44", "M.T.500", "M.T.7", "M.T.83", "M.T.8"),
  treatment = c(rep("control", 15), rep("treatment", 15))
  )

# running deseq normalization precursor

# Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData,
                              colData = metaData,
                              design = ~ treatment)

# Normalize the data
dds <- DESeq(dds)
```

### normalization

```{r}

dds <- estimateSizeFactors(dds)
sizeFactors(dds)
norm <- as.matrix(counts(dds, normalized=TRUE))
head(norm)

#saving as csv 
write.csv(norm, "../output/normalized_STAR.csv", row.names = T)

```

### need to bring back in gene names and rna type

```{r}

# importing norm as csv, because otherwise it stores as LHS or matrix (if specified). want as data frame for quicker processing
# plus, don't need to run the normalization / deseq steps if you just load normalized file
norm <- read.csv("../output/normalized_STAR.csv", row.names = 1)

# making dataframe of just genes and RNA type
types <- data.frame("gene" = rownames(counts), "Type" = counts$Type)
head(types)

# adding column to norm with gene names to prep for join
norm$gene <- rownames(norm)
head(norm)

# joining datframes to create new one with normalized count data and RNA type
norm_counts <- left_join(x = norm, y = types, by = "gene")
head(norm_counts)

# put genes back as row names
row.names(norm_counts) <- norm_counts$gene
head(norm_counts)

# fixing naming scheme for sample 193
colnames(norm_counts)[colnames(norm_counts) == 'M.T.193'] <- 'M.C.193'
head(norm_counts)
```

### T Tests

running t test for each RNA type with proportions. can change call from 'props' to 'counts' to run with raw counts if desired.

```{r}
# Identify control and treatment columns
control_cols <- c("M.C.193", "M.C.216", "M.C.218", "M.C.226", "M.C.306", "M.C.329", 
                  "M.C.334", "M.C.337", "M.C.339", "M.C.358", "M.C.360", "M.C.363", 
                  "M.C.373", "M.C.482", "M.C.488")
control_cols #peek

treatment_cols <- c("M.T.18", "M.T.20", "M.T.22", "M.T.235", "M.T.245", "M.T.31", 
                    "M.T.32", "M.T.399", "M.T.43", "M.T.44", "M.T.500", "M.T.7", 
                    "M.T.8", "M.T.83", "M.T.29")

treatment_cols  #peek
```

Separate dataframes for each gene type

```{r}
# specify data type here: proportion ('props'), raw counts ('counts'), or normalized counts ('norm_counts')
datatype <- props

# Subset the data for each gene type
mito_df <- datatype %>% filter(Type == "mito")
head(mito_df)

tRNA_df <- datatype %>% filter(Type == "tRNA")
head(tRNA_df)

protein_df <- datatype %>% filter(Type == "protein_coding")
head(protein_df)

lncRNA_df <- datatype %>% filter(Type == "lncRNA")
head(lncRNA_df)

rRNA_df <- datatype %>% filter(Type =="rRNA")
head(rRNA_df)

snRNA_df <- datatype %>% filter(Type =="snRNA")
head(snRNA_df)

snoRNA_df <- datatype %>% filter(Type =="snoRNA")
head(snoRNA_df)

ncRNA_df <- datatype %>% filter(Type =="ncRNA")
head(ncRNA_df)

pseudo_df <- datatype %>% filter(Type =="transcribed_pseudogene")
head(pseudo_df)
```

Format data frames to long, needed for t test

note that "values_to" is named as "count" — it is *proportion* of counts if 'props' is specified above, *normalized counts* if 'norm_counts' is specified above, or *raw counts* if 'counts' is specified aboe

```{r}
# Reshape data from wide to long format
mito_long <- mito_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(mito_long)

tRNA_long <- tRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(tRNA_long)

protein_long <- protein_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(protein_long)

lncRNA_long <- lncRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(lncRNA_long) 
        
rRNA_long <- rRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(rRNA_long)

snRNA_long <- snRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(snRNA_long)

snoRNA_long <- snoRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(snoRNA_long)

ncRNA_long <- ncRNA_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(ncRNA_long)

pseudo_long <- pseudo_df %>%
  pivot_longer(cols = all_of(c(control_cols, treatment_cols)),
               names_to = "group", values_to = "count") %>%
  mutate(condition = ifelse(group %in% control_cols, "control", "treatment"))
head(pseudo_long)
         
```

perform t tests

```{r}
mito_ttest <- t.test(count ~ condition, data = mito_long, var.equal = FALSE)
mito_ttest

tRNA_ttest <- t.test(count ~ condition, data = tRNA_long, var.equal = FALSE)
tRNA_ttest

protein_ttest <- t.test(count ~ condition, data = protein_long, var.equal = FALSE)
protein_ttest

lncRNA_ttest <- t.test(count ~ condition, data = lncRNA_long, var.equal = FALSE)
lncRNA_ttest

rRNA_ttest <- t.test(count ~ condition, data = rRNA_long, var.equal = FALSE)
rRNA_ttest

snRNA_ttest <- t.test(count ~ condition, data = snRNA_long, var.equal = FALSE)
snRNA_ttest

snoRNA_ttest <- t.test(count ~ condition, data = snoRNA_long, var.equal = FALSE)
snoRNA_ttest

ncRNA_ttest <- t.test(count ~ condition, data = ncRNA_long, var.equal = FALSE)
ncRNA_ttest 

pseudo_ttest <- t.test(count ~ condition, data = pseudo_long, var.equal = FALSE)
pseudo_ttest

```

t test results (uncorrected)

```{r}
# Collect results into a dataframe
ttest_results <- data.frame(
  Gene_Type = c("mito", "tRNA", "protein_coding", "lncRNA", "rRNA", 
                "snRNA", "snoRNA", "ncRNA", "pseudo"),
  p_value = c(mito_ttest$p.value, tRNA_ttest$p.value, protein_ttest$p.value,
              lncRNA_ttest$p.value, rRNA_ttest$p.value, snRNA_ttest$p.value,
              snoRNA_ttest$p.value, ncRNA_ttest$p.value, pseudo_ttest$p.value),
  t_statistic = c(mito_ttest$statistic, tRNA_ttest$statistic, protein_ttest$statistic,
                  lncRNA_ttest$statistic, rRNA_ttest$statistic, snRNA_ttest$statistic,
                  snoRNA_ttest$statistic, ncRNA_ttest$statistic, pseudo_ttest$statistic)
)

# View the t-test results dataframe
print(ttest_results)

```

testing corrections for repeated testing (needed here).

```{r}

# Apply Bonferroni correction
ttest_results$bonferroni_adjusted_p <- p.adjust(ttest_results$p_value, method = "bonferroni")

# Apply Benjamini-Hochberg correction (FDR)
ttest_results$fdr_adjusted_p <- p.adjust(ttest_results$p_value, method = "BH")

# View the results with adjusted p-values
print(ttest_results)

```

### Chi Square – Specify Data Type Here

Counts vs. Normalized Counts vs. Proportions

```{r}
# Summarize total proportions ('props') by gene type
# can use 'counts' to run by counts
# can use 'norm_counts' to run by normalized counts 

summary_table <- norm_counts %>%
  group_by(Type) %>%
  summarise(
    control_sum = sum(M.C.193, M.C.216, M.C.218, M.C.226, M.C.306, M.C.329, M.C.334, M.C.337, M.C.339, M.C.358, M.C.360, M.C.363, M.C.373, M.C.482, M.C.488),
    treatment_sum = sum(M.T.18, M.T.20, M.T.22, M.T.235, M.T.245, M.T.31, M.T.32, M.T.399, M.T.43, M.T.44, M.T.500, M.T.7, M.T.8, M.T.83, M.T.29)
  )

head(summary_table)
```

### Counts Chi Square

```{r}
# Convert to a matrix for statistical testing
contingency_table <- as.matrix(summary_table[, 2:3])
rownames(contingency_table) <- summary_table$Type

# Perform Chi-square test comparing control and treatment frequncies
chi_result <- chisq.test(contingency_table)
chi_result
```

### Frequencies Chi Square

```{r}

# the above table shows the frequencies for all samples, but since its totalling across 15 samples, the total frequency is 15, not 1. 
# here, I divide by 15 to create new dataframe where frequencies for control vs. treatment sums to 1

summary_table_new <- summary_table
summary_table_new[, -1] <- summary_table_new[, -1] / 15  # Divides all numeric columns by 15
summary_table_new

```

with frequencies for all 15 samples

```{r}
# Convert to a matrix for statistical testing
contingency_table <- as.matrix(summary_table[, 2:3])
rownames(contingency_table) <- summary_table$Type

# Perform Chi-square test comparing control and treatment frequncies
chi_result <- chisq.test(contingency_table)
chi_result

# Performing additonal chi square test where control frequencies are specified to be the "expected" frequencies
chi_res2 <- chisq.test(x = as.matrix(summary_table$treatment_sum), p = as.matrix(summary_table$control_sum), rescale.p = TRUE)
chi_res2 

```

Chi Square Test again but with frequencies divided by 15 (averaged across samples)

```{r}
# Convert to a matrix for statistical testing
contingency_table <- as.matrix(summary_table_new[, 2:3])
rownames(contingency_table) <- summary_table_new$Type

# Perform Chi-square test comparing control and treatment frequncies
chi_result <- chisq.test(contingency_table)
chi_result

# Performing additonal chi square test where control frequencies are specified to be the "expected" frequencies
chi_res2 <- chisq.test(x = as.matrix(summary_table_new$treatment_sum), p = as.matrix(summary_table_new$control_sum), rescale.p = TRUE)
chi_res2 
```

Chi Square results showed that there was a difference by gene type between control and treatment, but T test for each individual gene type (when FDR adjusted for multiple comparisons) revealed this was due to protein coding RNA types.

### Pie Chart

data prep

```{r}
pie_data <- props %>%
  group_by(Type) %>%
  summarise(
    control_sum = sum(M.C.193, M.C.216, M.C.218, M.C.226, M.C.306, M.C.329, M.C.334, M.C.337, M.C.339, M.C.358, M.C.360, M.C.363, M.C.373, M.C.482, M.C.488),
    treatment_sum = sum(M.T.18, M.T.20, M.T.22, M.T.235, M.T.245, M.T.31, M.T.32, M.T.399, M.T.43, M.T.44, M.T.500, M.T.7, M.T.8, M.T.83, M.T.29)
  )

pie_data[, -1] <- pie_data[, -1] / 15  # Divides all numeric columns by 15

head(pie_data)
```

chart itself

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# Convert data to long format
pie_long <- pie_data %>%
  pivot_longer(cols = c(control_sum, treatment_sum), 
               names_to = "group", 
               values_to = "value")

# Create the Control Pie Chart
control_pie <- ggplot(pie_long %>% filter(group == "control_sum"), 
                      aes(x = "", y = value, fill = factor(Type))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle("Control") +
  theme(legend.position = "right")

# Create the Treatment Pie Chart
treatment_pie <- ggplot(pie_long %>% filter(group == "treatment_sum"), 
                        aes(x = "", y = value, fill = factor(Type))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle("Treatment") +
  theme(legend.position = "right")

# Arrange the two plots side by side
control_pie + treatment_pie

```
